---
title: "Group Project"
author: "April Johns, Mollie Murphy, Josh Daniels"
date: "2025-11-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

pull in raw data 
```{r}
library(tidyverse)

groundhogs_raw <- read_csv('../data/groundhogs_raw.csv')
predictions_raw <- read_csv('../data/predictions_raw.csv')
avg_daily_temp_df <- read_csv('../data/average_daily_temp.csv')
```




Calculating the average temperature for each 6-week interval: 

For the preceding 6 weeks each year: 
- Groundhog Day is on February 2nd --> DOY = 33 (this col is in dataset, thankfully!)
- 6 weeks = 42 days 
- 42 - 33 = 9: we need the last 9 days of the preceding year 
- Therefore, we want to compute average temperature in the range: DOY-1: 357-366 (to include leap years), DOY: 1-33

For the 6 week interval in question:
- DOY between 34-75

For the following 6 weeks of spring: 
- DOY between 76-116 

```{r}
# data frame with average winter temp (6 weeks preceding groundhog day)
avg_winter_temp_df <- avg_daily_temp_df %>% 
  filter(DOY %in% c(1:33, 357:366)) %>% 
  mutate(groundhog_winter_year = if_else(DOY %in% c(357:366), YEAR + 1, YEAR)) %>% 
  group_by(id, groundhog_winter_year) %>% 
  summarise(avg_winter_temp = mean(T2M))

# data frame with average temp for prediction interval (6 weeks following groundhog day)
avg_prediction_temp_df <- avg_daily_temp_df %>% 
  filter(DOY %in% c(34:75)) %>% 
  group_by(id, YEAR) %>% 
  summarise(avg_prediction_temp = mean(T2M))

# data frame with average spring temp (6 weeks following prediction interval -- 6-12 weeks following groundhog day)
avg_spring_temp_df <- avg_daily_temp_df %>% 
  filter(DOY %in% c(75:116)) %>% 
  group_by(id, YEAR) %>% 
  summarise(avg_spring_temp = mean(T2M))
  
# join dataframes together, then compare which season prediction interval is closest to 
prediction_eval_df <- predictions_raw %>% 
  filter(year > 1980) %>% 
  left_join(avg_winter_temp_df, by = c('id', 'year' = 'groundhog_winter_year')) %>% 
  left_join(avg_prediction_temp_df, by = c('id', 'year' = 'YEAR')) %>% 
  left_join(avg_spring_temp_df, by = c('id', 'year' = 'YEAR')) %>% 
  mutate(delta_winter = abs(avg_winter_temp - avg_prediction_temp)) %>% 
  mutate(delta_spring = abs(avg_spring_temp - avg_prediction_temp)) %>% 
  mutate(six_more_weeks = delta_winter < delta_spring) %>% 
  mutate(is_correct = if_else(shadow & six_more_weeks, TRUE, FALSE))

```
